\vspace{-1.3cm} \textmd{\normalsize{High Performance Computing Project}}\\ Parallelization of the Probabilistic Roadmap Method with GPU Acceleration

### 2016-03-31 paper/paper.tex
...
rechtschreibung

Like in the stated robotics example, in many cases the most costly operation of the PRM algorithm are the many evaluations of the indicator function \(I(q)\), 
which however can be done independently for every new \(q\).
Therefore $I(q)$ is implemented in a vectorized way as a function \((q_1, ... , q_P) \mapsto (I(q_1),...,I(q_P))\) by a CUDA kernel, where every thread computes one evaluation.
In the robotics case, this means, the direct kinematic and collision calculation are made on the GPU parallely for different joint angle configurations.
Although this will the only major application in the project, the PRM solver and the configuration space are implemented seperated from each other and connected by an interface to be free to simply add other applications. As an example and for testing, additionally a simple 2D configuration space is provided, 
whose indicator function is read from a black-and-white image.

@@

While the distribution of the indicator function to different CUDA threads is a first level of parallelization, 
a second level can be achieved by sampling and connecting several new nodes parallely. 
This is more difficult, because all new nodes have to be connected 
and the graph data has to be updated on all processes. 
For this, different approaches have been implemented, which are described in chapter \ref{solver}.

@@

In the project, the probabilistic roadmap algorithm is applied  to create collisionfree trajectories of a robot arm.
For a robot with given joint angles \(q_i\) one can calculate straightforwardly all positions and orientations  of the single parts of the robot.
For this, coordinate frames according to the Denavit-Hartenberg convention are used.
The coordinate frames of the single parts are described by transformation matrices 
$T_i \in R^{4x4}$, which transform points from the body frame frame into the world frame. They depend on the state $q=(q_1,...,q_d)$ of the robot and on the geometry of the joint angles given by the Denavit-Hartenberg parameters of the robot.

@@

Now the movement of the robot can be described by a curve $q(t) \in \R^d$ in the configuration space. It is collisionfree, if at no time, any two polytopes of the robot intersect with each other or with the environment after being transformed into the world frame.
To apply the probabilistic roadmap algorithm, we express this by the indicator function

@@

The Denavit-Hartenberg (DH) convention is a method to describe the geometry of an robots joints, 
that is, how the transformations between the robots coordinate frames look like in dependancy of its joint values $q_1,...,q_d$.
For this it is assumed that each joint is either rotational, that means an rotation around a fixed axis, 
or prismatic, a translation along an axis.
The robot consists of $d+1$ bodies each of which has an body fixed coordinate frame $B_i$, which is described by a transformation matrix $T_i$ into the world frame.
For a robot with DH parameters $a_{i-1}, \alpha_{i-1}, d_i, \theta_i \in \R$, $i=1,...,d$, 
it is now defined, that the relative transformations $T_{i-1,i}=T_{i-1}^{-1} T_{i}$ between two bodies are
\begin{equation}
	T_{i-1,i} = R_z(\theta_i) T_z(d_i) R_x(\alpha_{i-1}) R_x(a_{i-1})
\end{equation}
where $R_x, R_z$ are rotations and $T_x, T_z$ translations around the $x$ and $z$ axes.
From this follows, that from given DH parameters all transformations are determined.
The other way round, one can show, that by an proper choice of the body frames, every robot with rotational or prismatic joints can be described by a set of DH parameters.
For the exact calculations, see \cite{robodyn}.

@@

To check if two polytopes collide an algorithm by Chung and Wang is used here. It uses the separating vector theorem and tries to find a seperating vector by iterative search. Furthermore, a subalgorithm determines in each iteration from the candidate history, if it is still theoretically possible to find a seperating vector. If not, it is returned a collision. If on the other side a seperating vector is found, the algorithm can return no collision.
For detailed informations see \cite{chung} and \cite{ericson}, pp. 410-412.
The method is implemented as device code for the use on GPU kernels.
\begin{lstlisting}
__host__ __device__ int seperating_vector_algorithm(const polytope4& P, const polytope4& Q, 
													const trafo4& tp, const trafo4& tq);
\end{lstlisting}
In our application, the polytopes are mostly parts of the robot, which are given in a body frame and have to be transformed into world frame according to section \ref{dk} first.
However, from the structure of the Chung-Wang algorithm drops out that mostly only a small part of the polytopes vertices are needed for the computations. Therefore it would be rather time wasting to transform all vertices before passing to the algorithm. To avoid this, all polytopes are passed in their body frames together with their transformations and the algorithm transforms vertices at the time they occur in calculations.

@@

\begin{algorithm}\label{kernel}
	\KwData{arrays qs, qe, testpos, testnum and threadindex}
	\KwResult{res}
	compute associated edge number $j$, defined by $testpos[j] \leq threadindex \le testpos[j]+testnum[j]$ \;
	$c = (threadindex - testpos[j]) / (testnum[k]-1) $ \;
	$q=c q_{start}^j + (1-c) q_{end}^j $ \;
	calculate robot kinematics $T_0(q), ..., T_d(q) $\;
	$res_{tmp} = 0; $ \;
	\For{all registered pairs pairs of polytopes $P, Q$}{
		$res_{tmp}=seperating\_vector\_algorithm(P, T_{i_P}, Q, T_{i_Q})$ \;
		\If{$res_{tmp} \neq 0$}{
			$res[j]=res_{tmp}$ \;
			break \;
		}
	}
	\caption{Robot kernel: code for one thread}
\end{algorithm}

In order to overlap computation time of GPU and CPU, an asynchronous version of the indicator function was written. Here the function is split in everything until the kernel launch and a wait function, which waits until the kernel finishes and recieves the result data.

@@

The implementation also allows overlapping of multiple requests, for example two calls of the launch function and after this the two corrsponding waiting calls. 
With this it is possible to use the full computation time of the GPU.

@@

At the end a shortest path from the start to the end node is searched by the Dijkstra algorithm.
It has shown that this part a neglectible amount of time compared to the graph building phase.
Therefore it is not parallelized here and the focus lies on building the roadmap.

@@

In the first approach, that was evaluated, every process generates new nodes independantly from each other.
Then the possible neighbours are determined and the connections, which have to be tested.
The indicator function is called and returnes the edges.
After this, the graph must be synchronized between all processes, which needs an all-to-all communication.
As the graph building phase does not need the edge data, it is sufficient to exchange only the nodes and collect the edges at the end.

@@

\subsubsection*{Version 2}
The second approach uses the fact, that by synchronizing the random seed, 
different processes can produce the same sequence of random numbers independently from each other.
By using this, every process can generate all new nodes and determine the connection data. 
Then the indicator function computation is splitted over their configuration space instances and the returned data shared afterwards with each other.
With this data, all graphs can be updated and it is guaranteed, that they grow exactly in the same way on each process.
The purpose of this version compared to the first one is, that less data has to be sent. 
Furthermore, in the first version it can occur, that the different generated nodes can result in very different counts of neighbors and therefore uneven loads on the GPUs.
Here, the distribution of the computation amount over the GPUs can be chosen almost freely.

@@

\begin{center}
\begin{tikzpicture}
	\tikzset{mywidth/.style={minimum width=2.5cm},};
	\node[block, mywidth] (l1) {\textbf{Worker 1}};
	\node[frameless, mywidth, below=of l1, minimum height=1.5cm] (l2) {$\vdots$};
	\node[block, mywidth, below=of l2] (l3) 	{\begin{tabular}{c}
										generate\\ nodes, \\
										start kernel
									\end{tabular}};
	\node[frameless, mywidth, below=of l3, minimum height=2cm] (l6) {};
	\node[frameless, mywidth, below=of l6] (l7) {};
	\node[block, mywidth, below=of l7] (l11)	{\begin{tabular}{c}
										recieve\\ result, \\
										store nodes \\ 
										and edges
									\end{tabular}};
	\node[frameless, mywidth, below=of l11] (l12) {$\vdots$};

	\node[block, mywidth, right=of l1] (r1) {\textbf{Worker 2}};
	\node[frameless, mywidth, right=of l2, minimum height=1.5cm] (r2) {$\vdots$};
	\node[frameless, mywidth, right=of l3] (r3) {};
	\node[block, mywidth, right=of l6] (r6)	{\begin{tabular}{c}
										recieve\\ result, \\
										store nodes \\
										and edges
									\end{tabular}};
	\node[block, mywidth, right=of l7] (r7) 	{\begin{tabular}{c}
										generate\\ nodes, \\
										start kernel
									\end{tabular}};
	\node[frameless, mywidth, right=of l12] (r12) {$\vdots$};

@@

\begin{wrapfigure}{r}{0.5\textwidth}\label{robotsolution}
	\centering
	\includegraphics[width=0.5\textwidth]{robotsolution.pdf}
	\caption{PRM generated collisionfree movement for a 4-axis robot}
\end{wrapfigure}

Tests have been made for a scenario with a robot with 4 joints. For creating and plotting the environment and the resulting trajectories, Matlab functions are provided.
Because of the dependancy on random numbers, the algorithm takes different amounts of time for different runs.
Therefore, 10 runs were made for each version and number of processes.
However, the different versions lead to a different average amount of GPU threads needed.
So, to be able to better compare how they scale, the time per number of GPU threads used as a quantity.
Table \ref{resulttable} shows the results.
One can see, that while the versions with node communication has a better time to GPU threads ratio, but leads to a greater amount of threads on average.
The versions with less communication however have better absolute times.
Furthermore, one sees the improvement from version 2 to version 3 from the asynchronous indicator function.

@@

\end{tabular}
\caption{Timing results for the parallelization versions from chapter \ref{solver} on a machine with 4 GeForce GTX 480}
\end{table}

>>>>

Like in the stated robotics example, in many cases the most costly operation of the PRM algorithm are the many evaluations of the indicator function \(I(q)\), 
which however can be done independently for every new \(q\).
Therefore $I(q)$ is implemented in a vectorized way as a function \((q_1, ... , q_P) \mapsto (I(q_1),...,I(q_P))\) by a CUDA kernel, where every thread computes one evaluation.
In the robotics case, this means, the direct kinematic and collision calculation are made on the GPU parallelly for different joint angle configurations.
Although this will the only major application in the project, the PRM solver and the configuration space are implemented separated from each other and connected by an interface to be free to simply add other applications. As an example and for testing, additionally a simple 2D configuration space is provided, 
whose indicator function is read from a black-and-white image.

@@

While the distribution of the indicator function to different CUDA threads is a first level of parallelization, 
a second level can be achieved by sampling and connecting several new nodes parallelly. 
This is more difficult, because all new nodes have to be connected 
and the graph data has to be updated on all processes. 
For this, different approaches have been implemented, which are described in chapter \ref{solver}.

@@

In the project, the probabilistic roadmap algorithm is applied  to create collision-free trajectories of a robot arm.
For a robot with given joint angles \(q_i\) one can calculate straightforwardly all positions and orientations  of the single parts of the robot.
For this, coordinate frames according to the Denavit-Hartenberg convention are used.
The coordinate frames of the single parts are described by transformation matrices 
$T_i \in R^{4x4}$, which transform points from the body frame frame into the world frame. They depend on the state $q=(q_1,...,q_d)$ of the robot and on the geometry of the joint angles given by the Denavit-Hartenberg parameters of the robot.

@@

Now the movement of the robot can be described by a curve $q(t) \in \R^d$ in the configuration space. It is collision-free, if at no time, any two polytopes of the robot intersect with each other or with the environment after being transformed into the world frame.
To apply the probabilistic roadmap algorithm, we express this by the indicator function

@@

The Denavit-Hartenberg (DH) convention is a method to describe the geometry of an robots joints, 
that is, how the transformations between the robots coordinate frames look like in dependency of its joint values $q_1,...,q_d$.
For this it is assumed that each joint is either rotational, that means an rotation around a fixed axis, 
or prismatic, a translation along an axis.
The robot consists of $d+1$ bodies each of which has an body fixed coordinate frame $B_i$, which is described by a transformation matrix $T_i$ into the world frame.
For a robot with DH parameters $a_{i-1}, \alpha_{i-1}, d_i, \theta_i \in \R$, $i=1,...,d$, 
it is now defined, that the relative transformations $T_{i-1,i}=T_{i-1}^{-1} T_{i}$ between two bodies are
\begin{equation}
	T_{i-1,i} = R_z(\theta_i) T_z(d_i) R_x(\alpha_{i-1}) R_x(a_{i-1})
\end{equation}
where $R_x, R_z$ are rotations and $T_x, T_z$ translations around the $x$ and $z$ axes.
From this follows, that from given DH parameters all transformations are determined.
The other way round, one can show, that by an proper choice of the body frames, every robot with rotational or prismatic joints can be described by a set of DH parameters.
For the exact calculations, see \cite{robodyn}.

@@

To check if two polytopes collide an algorithm by Chung and Wang is used here. It uses the separating vector theorem and tries to find a separating vector by iterative search. Furthermore, a sub-algorithm determines in each iteration from the candidate history, if it is still theoretically possible to find a separating vector. If not, it is returned a collision. If on the other side a separating vector is found, the algorithm can return no collision.
For detailed informations see \cite{chung} and \cite{ericson}, pp. 410-412.
The method is implemented as device code for the use on GPU kernels.
\begin{lstlisting}
__host__ __device__ int separating_vector_algorithm(const polytope4& P, const polytope4& Q, 
													const trafo4& tp, const trafo4& tq);
\end{lstlisting}
In our application, the polytopes are mostly parts of the robot, which are given in a body frame and have to be transformed into world frame according to section \ref{dk} first.
However, from the structure of the Chung-Wang algorithm drops out that mostly only a small part of the polytopes vertices are needed for the computations. Therefore it would be rather time wasting to transform all vertices before passing to the algorithm. To avoid this, all polytopes are passed in their body frames together with their transformations and the algorithm transforms vertices at the time they occur in calculations.

@@

\begin{algorithm}\label{kernel}
	\KwData{arrays qs, qe, testpos, testnum and threadindex}
	\KwResult{res}
	compute associated edge number $j$, defined by $testpos[j] \leq threadindex \le testpos[j]+testnum[j]$ \;
	$c = (threadindex - testpos[j]) / (testnum[k]-1) $ \;
	$q=c q_{start}^j + (1-c) q_{end}^j $ \;
	calculate robot kinematics $T_0(q), ..., T_d(q) $\;
	$res_{tmp} = 0; $ \;
	\For{all registered pairs pairs of polytopes $P, Q$}{
		$res_{tmp}=separating\_vector\_algorithm(P, T_{i_P}, Q, T_{i_Q})$ \;
		\If{$res_{tmp} \neq 0$}{
			$res[j]=res_{tmp}$ \;
			break \;
		}
	}
	\caption{Robot kernel: code for one thread}
\end{algorithm}

In order to overlap computation time of GPU and CPU, an asynchronous version of the indicator function was written. Here the function is split in everything until the kernel launch and a wait function, which waits until the kernel finishes and receives the result data.

@@

The implementation also allows overlapping of multiple requests, for example two calls of the launch function and after this the two corresponding waiting calls. 
With this it is possible to use the full computation time of the GPU.

@@

At the end a shortest path from the start to the end node is searched by the Dijkstra algorithm.
It has shown that this part a neglectable amount of time compared to the graph building phase.
Therefore it is not parallelized here and the focus lies on building the roadmap.

@@

In the first approach, that was evaluated, every process generates new nodes independently from each other.
Then the possible neighbors are determined and the connections, which have to be tested.
The indicator function is called and returns the edges.
After this, the graph must be synchronized between all processes, which needs an all-to-all communication.
As the graph building phase does not need the edge data, it is sufficient to exchange only the nodes and collect the edges at the end.

@@

\subsubsection*{Version 2}
The second approach uses the fact, that by synchronizing the random seed, 
different processes can produce the same sequence of random numbers independently from each other.
By using this, every process can generate all new nodes and determine the connection data. 
Then the indicator function computation is split over their configuration space instances and the returned data shared afterwards with each other.
With this data, all graphs can be updated and it is guaranteed, that they grow exactly in the same way on each process.
The purpose of this version compared to the first one is, that less data has to be sent. 
Furthermore, in the first version it can occur, that the different generated nodes can result in very different counts of neighbors and therefore uneven loads on the GPUs.
Here, the distribution of the computation amount over the GPUs can be chosen almost freely.

@@

\begin{center}
\begin{tikzpicture}
	\tikzset{mywidth/.style={minimum width=2.5cm},};
	\node[block, mywidth] (l1) {\textbf{Worker 1}};
	\node[frameless, mywidth, below=of l1, minimum height=1.5cm] (l2) {$\vdots$};
	\node[block, mywidth, below=of l2] (l3) 	{\begin{tabular}{c}
										generate\\ nodes, \\
										start kernel
									\end{tabular}};
	\node[frameless, mywidth, below=of l3, minimum height=2cm] (l6) {};
	\node[frameless, mywidth, below=of l6] (l7) {};
	\node[block, mywidth, below=of l7] (l11)	{\begin{tabular}{c}
										receive\\ result, \\
										store nodes \\ 
										and edges
									\end{tabular}};
	\node[frameless, mywidth, below=of l11] (l12) {$\vdots$};

	\node[block, mywidth, right=of l1] (r1) {\textbf{Worker 2}};
	\node[frameless, mywidth, right=of l2, minimum height=1.5cm] (r2) {$\vdots$};
	\node[frameless, mywidth, right=of l3] (r3) {};
	\node[block, mywidth, right=of l6] (r6)	{\begin{tabular}{c}
										receive\\ result, \\
										store nodes \\
										and edges
									\end{tabular}};
	\node[block, mywidth, right=of l7] (r7) 	{\begin{tabular}{c}
										generate\\ nodes, \\
										start kernel
									\end{tabular}};
	\node[frameless, mywidth, right=of l12] (r12) {$\vdots$};

@@

\begin{wrapfigure}{r}{0.5\textwidth}\label{robotsolution}
	\centering
	\includegraphics[width=0.5\textwidth]{robotsolution.pdf}
	\caption{PRM generated collision-free movement for a 4-axis robot}
\end{wrapfigure}

Tests have been made for a scenario with a robot with 4 joints. For creating and plotting the environment and the resulting trajectories, Matlab functions are provided.
Because of the dependency on random numbers, the algorithm takes different amounts of time for different runs.
Therefore, 10 runs were made for each version and number of processes.
However, the different versions lead to a different average amount of GPU threads needed.
So, to be able to better compare how they scale, the time per number of GPU threads used as a quantity.
Table \ref{resulttable} shows the results.
One can see, that while the versions with node communication has a better time to GPU threads ratio, but leads to a greater amount of threads on average.
The versions with less communication however have better absolute times.
Furthermore, one sees the improvement from version 2 to version 3 from the asynchronous indicator function.

@@

\end{tabular}
\caption{Timing results for the parallelization versions from chapter \ref{solver} on a cluster node with 2x Intel Xeon Six-Core CPU X5650 @2.67GHz and 4x GeForce GTX 480}
\end{table}
